A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-06-22 example.tex
WIP: Rename the Cut predicate and LazyParikh!
Closes: #51
Closes: #14
Tweak figures
Much tighter integration between sections

To solve these constraints, we will first perform translation into Parikh
automata, then break down how the constraints will be handled by \Calculus{}.

@@

By a Parikh automaton we mean a standard non-deterministic finite automaton
(NFA) with a set of integer registers (Parikh registers) that are incremented
(or decremented) at each transition.

To solve \cref{ex:string-constraints} using \Calculus{} in a theorem proving
context, \cref{const:s1-in-c-dd,const:s2-in-b,const:s1-substring} can be
translated to the product of Parikh automata $B \times A'$, where $A'$
(\cref{fig:aut_a}) is the Parikh automata obtained by applying the encoding of
$\Substring$ described in \cite{ostrich-plus}. That description can then be used
to obtain values for the subsstring offsets $i, n$. The corresponding automata
can be found in \cref{fig:examples}.

@@

\begin{figure}[ht]
    \centering 
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{a}
    \caption{The automaton $A'$, whose Parikh registers count the length of the
    string accepted by the automaton (register $l_A$), the start offset of the
    substring ($i$) and length ($n$) of the substring matching
    $\Regex{c\RegexOr{}dd}$. Note the symbolic transitions in the starting and
    accepting states matching any symbol in the alphabet!}\label{fig:aut_a}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{b}
    \caption{The automaton $B$, where the Parikh registers count the length of
    the string in register $l_B$.}\label{fig:aut_b}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The collection of automata we use as running
  examples, both derived from \cref{ex:string-constraints}.}\label{fig:examples}
\end{figure}

@@

\begin{figure}[h]
  \centering 
  \includegraphics[scale=\autscale]{a_1}
  \caption{ $A'$ with its associated transition variables in symbolic form after
  minimal elminiation of variables using linear algebra. Note the incoming flow
  of $1$ in the starting state and its balancing of the outgoing transitions!
  }\label{fig:a_1}
\end{figure}

@@

\begin{figure}[h]
  \centering 
  \includegraphics[scale=\autscale]{a_2}
  \caption{ $A'$ after even more linear algebra, with most transitions expressed
  in terms of $n$. Intuitively, this version captures the fact that $n$ is given
  by distributing the incoming flow of $1$ across the two outgoing transitions
  from $I_{A'}$, the initial state.}
  \label{fig:a_2}
\end{figure}

@@

\begin{figure}[h]
    \centering 
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{a_annotated}
    \caption{ $A'$ with its associated transition variables in symbolic
    form.}\label{fig:aut_a_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \begin{subfigure}[b]{0.5\textwidth}
    \centering
    \includegraphics[scale=\autscale]{b_annotated}
    \caption{$B$ with its associated transition variables in symbolic forms.
    Note the large number of implicitly existence-quantified bound variables on
    transitions, suggesting that this automaton has a more complex structure
    with relation to its (free) target variable representing the string length
    than $A'$. The variables used here are fresh, and do not correspond to the
    ones previously used.}\label{fig:aut_b_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The automata after applying linear algebra and equality elimination
  to prune them.}\label{fig:propagated}
\end{figure}

>>>>

To solve these constraints, we will first perform translation into Parikh
automata, then break down how the constraints will be handled by \Calculus{}.
This example will recur in a more formal fashion in \cref{sec:generalised},
which can be read in parallel if (even) more detail is desired.

@@

By a Parikh automaton we mean a standard non-deterministic finite automaton
(NFA) with a set of integer registers (Parikh registers) that are incremented
(or decremented) at each transition. A formal definition can be found in
\cref{def:parikh-automata}.

To solve \cref{ex:string-constraints} using \Calculus{} in a theorem proving
context, \cref{const:s1-in-c-dd,const:s2-in-b,const:s1-substring} can be
translated to the product of Parikh automata $B \times A'$, where $A'$
(\cref{fig:aut_a}) is the Parikh automata obtained by applying the encoding of
$\Substring$ described in \cite{ostrich-plus} to
$\Language(\Regex{c\RegexOr{}dd})$. That description can then be used to obtain
values for the substring offsets $i, n$. The corresponding automata can be found
in \cref{fig:examples}.

@@

\newcommand{\autscale}[0]{0.48}

\begin{figure}[ht]
    \centering 
  \begin{subfigure}[b]{\autscale\textwidth}
    \centering
    \includegraphics[width=\textwidth]{a}
    \caption{The automaton $A'$, whose Parikh registers count the length of the
    string accepted by the automaton (register $l_A$), the start offset of the
    substring ($i$) and length ($n$) of the substring matching
    $\Regex{c\RegexOr{}dd}$. Note the symbolic transitions in the starting and
    accepting states matching any symbol in the alphabet!}\label{fig:aut_a}
      %\Description[SHORT]{LONG}
  \end{subfigure}%
  \hfill%
  \begin{subfigure}[b]{\autscale\textwidth}
    \centering
    \includegraphics[width=\textwidth]{b}
    \caption{The automaton $B$, where the Parikh registers count the length of
    the string in register $l_B$.}\label{fig:aut_b}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The collection of automata we use as running
  examples, both derived from \cref{ex:string-constraints}.}\label{fig:examples}
\end{figure}

@@

\begin{figure}[h]
  \centering 
  \includegraphics[width=0.6\textwidth]{a_1}
  \caption{ $A'$ with its associated transition variables in symbolic form after
  minimal elminiation of variables using linear algebra. Note the incoming flow
  of $1$ in the starting state and its balancing of the outgoing transitions!
  }\label{fig:a_1}
\end{figure}

@@

\begin{figure}[h]
  \centering 
  \includegraphics[width=0.6\textwidth]{a_2}
  \caption{ $A'$ after even more linear algebra, with most transitions expressed
  in terms of $n$. Intuitively, this version captures the fact that $n$ is given
  by distributing the incoming flow of $1$ across the two outgoing transitions
  from $I_{A'}$, the initial state.}
  \label{fig:a_2}
\end{figure}

@@

\begin{figure}[h]
    \centering 
  \begin{subfigure}[b]{\autscale\textwidth}
    \centering
    \includegraphics[width=\textwidth]{a_annotated}
    \caption{ $A'$ with its associated transition variables in symbolic
    form.}\label{fig:aut_a_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}\hfill%
  \begin{subfigure}[b]{\autscale\textwidth}
    \centering
    \includegraphics[width=\textwidth]{b_annotated}
    \caption{$B$ with its associated transition variables in symbolic forms.
    Note the large number of implicitly existence-quantified bound variables on
    transitions, suggesting that this automaton has a more complex structure
    with relation to its (free) target variable representing the string length
    than $A'$. The variables used here are fresh, and do not correspond to the
    ones previously used.}\label{fig:aut_b_annotated}
      %\Description[SHORT]{LONG}
  \end{subfigure}
  \caption{The automata after applying linear algebra and equality elimination
  to prune them.}\label{fig:propagated}
\end{figure}

### 2023-06-22 generalised.tex
Much tighter integration between sections

\subsection{Integer Datatype Constraints on Strings}\label{sec:parikh-automata}

Parikh images are also applicable for deciding more general classes of
string constraints~\cite{ostrich-plus}. Consider, for instance, a
constraint involving the substring operation of \cref{const:s1-substring}:
\begin{equation}\label{eq:substring}
    s_1 = \Substring(s_2, i, n)
\end{equation}

That is $s_1$ is an $n$-length substring of $s_2$ starting at offset $i$ (in
some variants , $n$ is $i + |n|$, i.e. the ending index). That constraint
belongs to an expressive fragment of string logic that cannot be decided by most
state-of-the-art string solvers.

We can model \cref{eq:substring} using the notion of a \emph{Parikh
  automaton}~\cite{parikh-automata,expressiveness}. A Parikh automaton
is a standard automaton is extended such that transitions are labelled both with characters
from the alphabet~$\Alphabet$, and with offset vectors defining the
increments of a finite number of counters. This means that Parikh
automata recognise words over an extended
alphabet~$\Alphabet \times D$, where $D \subseteq \Naturals^d$ is a
finite set of increment vectors (notation as
in~\cite{expressiveness}). We use the symbols~$\pi_\Alphabet, \pi_D$
to denote projections to the first and the second component of a
composite letter~$(a, \Vector{v})$, respectively, and extend those
projections to words:
\begin{equation*}
  \pi_\Alphabet((a_1, \Vector{v}_1) \RepeatSum{\Concat} (a_k, \Vector{v}_k))
  ~=~ a_1 \RepeatSum{\Concat} a_k,
  \qquad
  \pi_D((a_1, \Vector{v}_1) \RepeatSum{\Concat} (a_k, \Vector{v}_k))
  ~=~ \Vector{v}_1 \RepeatSum{+} \Vector{v}_k.
\end{equation*}

@@

    Applied to \eqref{eq:substring}, the decision procedure in
\cite{ostrich-plus} will construct a pre-image of $\Language_2$ under
the substring operation~$\cdot[n:m]$, and check whether this pre-image
is consistent with the constraint~$x \in \Language_1$. Because the
substring operation depends on the values of the integer
variables~$n, m$, a Parikh automaton of dimension~$2$ is a suitable
formalism to describe the pre-image, resulting
in the automaton in Fig.~\ref{fig:parikh-automata} (right). The first
component of the increment vectors is used to count the number of
letters eliminated in the beginning of the string (value~$n$), while
the second component records the beginning of the eliminated suffix
(value~$m$).

Denoting the language described by Fig.~\ref{fig:parikh-automata}
(right) as $\Language_{pre}$, we can then replace \eqref{eq:substring}
with an equi-satisfiable formula that no longer contains any explicit
substring operation:
%
\begin{equation}
  \label{eq:parikh-constraint}
  p \in \Language_{pre} \wedge~
  \pi_\Alphabet(p) \in \Language_1
  \wedge~~ \pi_D(p) =
  \begin{bmatrix}
    n \\ m
  \end{bmatrix}~~
  \wedge 0 \leq n \leq m \leq |p|
\end{equation}
%
To check the satisfiability of \eqref{eq:parikh-constraint}, we need a
decision procedure that can process intersections of regular languages
(in this case, of $\Language_{pre}$ and $\Language_1$, synchronising
on $\Alphabet$), while imposing the side
condition~$0 \leq n \leq m \leq |p|$ on the increment sum. In
\cite{ostrich-plus}, this decision procedure turned out to be main
bottleneck of the string solver, which was one of the motivations to
develop the lazy algorithm proposed in this paper.

\iffalse
\begin{figure}[t]
  \centering
      \includegraphics[width=0.5\linewidth]{parikh_automaton}
      \caption{The automaton part of a Parikh automaton for $\AcaOrBc{}$ with
      $\Alphabet = \Set{\text{a, b, c}}, d = 3$. The semilinear set/Presburger
      formula containing the constraints on the final register values cannot be
      visualised.}\label{fig:parikh-automaton}
    \end{figure}
\fi

>>>>

\subsection{Integer Constraints on Strings}\label{sec:parikh-automata}

Parikh images are also applicable for deciding more general classes of string
constraints~\cite{ostrich-plus}. Consider the substring constraint
\cref{const:s1-substring} of \cref{ex:string-constraints}: $ s_1 =
\Substring(s_2, i, n)$, that is $s_1$ is an $n$-length substring of $s_2$
starting at offset $i$ (in some variants , $n$ is $i + |n|$, i.e. the ending
index). That constraint belongs to an expressive fragment of string logic that
cannot be decided by most state-of-the-art string solvers.

In \cref{sec:intuition} we hand-wavily modelled \cref{const:s1-substring} (and
  the other constraints) using \emph{Parikh
  automata}~\cite{parikh-automata,expressiveness}. A Parikh automaton is a
  standard NFA extended such that transitions are additionally labelled both
  with offset vectors defining the increments of a finite number of counters.
  This means that Parikh automata recognise words over an extended
  alphabet~$\Alphabet \times D$, where $D \subseteq \Naturals^d$ is a finite set
  of increment vectors (notation as in~\cite{expressiveness}) and $\Alphabet$ is
  the alphabet of the original NFA.

  We use the symbols~$\pi_\Alphabet, \pi_D$ to denote projections to the first
  and the second component of a composite letter~$(a, \Vector{v})$,
  respectively, and extend those projections to words:
\begin{equation*}
  \pi_\Alphabet((a_1, \Vector{v}_1) \RepeatSum{\Concat} (a_k, \Vector{v}_k))
  ~=~ a_1 \RepeatSum{\Concat} a_k,
  \qquad
  \pi_D((a_1, \Vector{v}_1) \RepeatSum{\Concat} (a_k, \Vector{v}_k))
  ~=~ \Vector{v}_1 \RepeatSum{+} \Vector{v}_k.
\end{equation*}

@@

Applied to \cref{const:s1-substring}, the decision procedure in
\cite{ostrich-plus} will construct a pre-image of $s_1$ under the substring
operation, and check whether this pre-image is consistent with the constraint~$x
\in \Language_1$, corresponding to the full product seen in
\cref{ex:string-constraints}. Because the substring operation depends on the
values of the integer variables~$i, n$, a Parikh automaton of dimension~$2$,
shown in \cref{fig:aut_a}, models the pre-image. Intuitively, this construction
accumulates any symbol (for the prefix of the substring operation), incrementing
$i$ to mark the start of the substring, followed by the automaton representing
the substring itself, modified to add a counter to increment $n$ at each
transition, followed by a state that performs no increments and only accepts any
suffix after the matched substring.

% @Philipp: I am quite unsure about which intersection we are looking for here,
% please verify I'm not wrong!
Denoting the language described by \cref{fig:aut_a} as $\Language_{pre}$, we can
then replace $s_1 = \Substring(s_2, i, n)$ and $s_2 \in
\Language(B)$ (of \cref{const:s1-in-c-dd}) with an
equi-satisfiable formula that no longer contains any explicit substring
operation:
%
\begin{equation}
  \label{eq:parikh-constraint}
  p \in \Language_{pre} \wedge~
  \pi_\Alphabet(p) \in \Language(B)
  \wedge~~ \pi_D(p) =
  \begin{bmatrix}
    i \\ n
  \end{bmatrix}~~
  \wedge 0 \leq i \leq i + n \leq |p|
\end{equation}
%
To check the satisfiability of \cref{eq:parikh-constraint}, we need a
decision procedure that can process intersections of regular languages
(in this case, of $\Language_{pre}$ and $\Language(B)$, synchronising
on $\Alphabet$), while imposing the side
condition~$0 \leq n \leq m \leq |p|$ on the increment sum. In
\cite{ostrich-plus}, this decision procedure turned out to be main
bottleneck of the string solver, which was one of the motivations to
develop the lazy algorithm proposed in this paper.
